Linear analysis of the vertical shear instability in astrophysical disks

### 2015-02-09 appendix.tex
major updating

\section{Applicability of the vertically global shearing box}\label{global_corr}
In \S\ref{analytical} we used the vertically global shearing box
\citep[VGSB,][]{mcnally14} to make analytical progress for vertically
isothermal disks. Here, we discuss limitations of this approach.

The main difference between  VGSB framework

Using the notation and disk models in the main text, the
energy, density and $r$ momentum equations become 
\begin{align}
  \ii \sigma W  &=c _s^2(r_0, z)\left( \left.\frac{\p\ln\rho}{\p
      z}\right|_0 \delta v_z + \frac{\gamma}{\Gamma} \frac{d\delta
    v_z}{dz}\right) + \left[\ii k_rc_s^2(r_0,z)
  \frac{\gamma}{\Gamma} + \hat{g}_c\left.\frac{1}{\rho}\frac{\p P}{\p
      r}\right|_0\right]\delta v_r  +
\frac{1}{t_c}\left(W-\frac{Q}{\Gamma}\right),\label{gcorr_terms1}\\
\ii \sigma Q &= c_s^2(r_0,z)\left(\ii k_r + \hat{g}_c
  \left.\frac{\p\ln{\rho}}{\p r}\right|_0\right)\delta v_r + c _s^2(r_0, z)\left( \left.\frac{\p\ln\rho}{\p
      z}\right|_0 \delta v_z + \frac{d\delta
    v_z}{dz}\right),\label{gcorr_terms2} \\
\ii \sigma \delta v_r & = \ii k_r W  -
\hat{g}_c\frac{1}{c_s^2(r_0,z)}\left.\frac{1}{\rho}\frac{\p P}{\p
  r}\right|_0Q - 2\Omega(r_0,z)\delta v_\phi,\label{gcorr_terms3}
\end{align}
where subscript $0$ denotes evaluation at $r=r_0$.

@@

\subsection{Spurious growth of adiabatic perturbations in the VGSB}\label{analytic_adia} 
For adiabatic perturbations ($\gamma>1$, $\beta\to\infty$) we may
simply use the Soilberg-Hoiland critera to assess stability
(\S\ref{solberg}). Nevertheless, we briefly consider this case in the
local framework to demonstrate its limitations. For $\beta\to\infty$
the parameter $\chi = 1/\gamma$ and Eq. \ref{vertiso_gov_nondim} can
be written as

\begin{align}
  0 =\dd v_z^{\prime\prime} + \left(1 + \ii \epsilon q
    \hat{k}\right)\left(\ln\rho^{\prime}\delta v_z\right)^\prime
  +\left\{\hat{\sigma}^2\left(\frac{1}{\gamma}+\hat{k}^2\right) 
    -\left(\frac{\gamma-1}{\gamma}\right)\left[\ln\rho^{\prime\prime}+\hat{k}^2\left(1-\frac{\ii\epsilon  
          q}{\hat{k}}\right)\ln\rho^{\prime 2}\right]\right\}\delta v_z.\label{adia_iso3}
\end{align}

We multiply Eq. \ref{adia_iso3} by $\rho\delta v_z^*$ and
integrate vertically, assume boundary terms vanish when integrating by
parts, to obtain
\begin{align}
  \hat{\sigma}^2\left(\frac{1}{\gamma} +
    \hat{k}^2\right)\int_{\zhat_1}^{\zhat_2}\rho|\delta
  v_z|^2 d\zhat 
  =&  \left(\frac{\gamma-1}{\gamma}\right)
  \int_{\zhat_1}^{\zhat_2}\rho|\delta v_z^\prime|^2 d\zhat
  +\frac{1}{\gamma}\int_{\zhat_1}^{\zhat_2}\frac{1}{\rho}|(\rho\delta
  v_z)^\prime|^2 d\zhat\notag\\
&+
  \left(\frac{\gamma-1}{\gamma}\right)\hat{k}^2\left(1-\frac{\ii\epsilon
      q}{\hat{k}}\right) \int_{\zhat_1}^{\zhat_2}\rho\ln\rho^{\prime
    2}|\delta v_z|^2 d\zhat
+ \ii\epsilon q \hat{k}
  \int_{\zhat_1}^{\zhat_2}\ln\rho^\prime(\rho\delta v_z^*)^\prime
  \delta v_z d\zhat.\label{adia_integral}
\end{align}
When $q\equiv0$, all the terms on the right-hand-side (RHS) are real. Then
$\hat{\sigma}^2$ is real, and  $\hat{\sigma}^2>0$ if $\gamma>1$. As
expected, a sub-adiabatically stratified disk is stable in the absense
of vertical shear.

However, in the presence of vertical shear $q\neq0$,
Eq. \ref{adia_integral} suggests $\hat{\sigma}$ is complex for real
$\khat$, which implies the possibility of instability, regardless of
how much $\gamma$ deviates from unity. This is contradicts the second 
Solberg-Hoiland criterion (Eq. \ref{solberg2}) and is a result of the
radially-local approximation (Appendix \ref{global_corr}). To 
avoid this, we must restrict our attention to short thermal relaxation
timescales ($\beta\lesssim1$) and/or neutrally-stratified disks
($\gamma=\Gamma$) when $q\neq0$.

\subsection{Effect of global radial gradients}
We may also reverse the above procedure by reinstating the `global
correction' terms proportional to $\hat{g}_c$ in the linear VGSB
equations. Of course, doing so implies the basic state is not uniform
in $x$, which would be inconsistent with the full VGSB equations in
equilibrium. Nevertheless, we illustrate in Fig. \ref{gcorr_compare} the effect of
these global radial gradients on the growth rate of the fundamental
VSI.

The global correction terms do not change the qualitative dependence
of the fundamental VSI growth rates on the thermal relaxation
timescale. The difference is negligble for $\beta\lesssim 1$ (i.e.
fast thermal relaxation). The critical relaxation timescale for the
fundamental VSI is unaffected by these global radial radients.

However, in the adiabatic limit $\beta\to\infty$ the linear VGSB
equations give a positive growth rate as discussed in
\S\ref{analytic_adia}, which is inconsistent with the second
Solberg-Hoiland criterion (Eq. \ref{solberg2}). The correct behaviour
($\nu\to0$ as $\beta\to\infty$) is obtained with global correction.

It is interesting to note that there is a thermal relaxation timescale
that maximizes the mode decay rate. Here, it is $\beta\simeq2$ or
about $0.3$ orbital periods. This is, in fact, consistent with Fig. 12
of \cite{nelson13}.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.0cm 0cm
  0cm]{figures/gcorr_compare} 
  \caption{Growth rate of
    the fundamental VSI mode as a function of the thermal
    relaxatin time $\beta$, computed from the linear VGSB equations used in
    the main text (dotted) and that including additional global
    radial gradient terms in  Eq. \ref{gcorr_terms1}---\ref{gcorr_terms3} (solid). 
    The disk parameters are $\Gamma=1.011$,
    $\gamma=1.4$ and $(p,q,\epsilon)=(-1.5,-1,0.05)$. The perturbation
    wavenumber is $\khat=20$. 
    {\bf waiting for
      high resolution} 
    \label{gcorr_compare}}  
\end{figure}

>>>>

\section{Applicability of the vertically global shearing box}\label{global_corr}
In \S\ref{analytical} we used the vertically global shearing box
\citep[VGSB,][]{mcnally14} to make analytical progress. % Here, we discuss limitations of this approach.   
The main difference between the linear problem in the VGSB 
and the radially-local approximation
(Eq. \ref{lin_mass}---\ref{lin_energy}) adopted in our numerical study
is the absence of  terms associated with the radial disk structure in
the former framework. These terms appear in the energy, mass and radial momentum
equations and are reproduced here for convenience,
% Using the notation and disk models in the main text, the
% energy, density and $r$ momentum equations become 
\begin{align}
  \ii \sigma W  &=c _s^2(r_0, z)\left( \left.\frac{\p\ln\rho}{\p
      z}\right|_0 \delta v_z + \frac{\gamma}{\Gamma} \frac{d\delta
    v_z}{dz}\right) + \left[\ii k_xc_s^2(r_0,z)
  \frac{\gamma}{\Gamma} + \hat{g}_c\left.\frac{1}{\rho}\frac{\p P}{\p
      r}\right|_0\right]\delta v_x  +
\frac{1}{t_c}\left(W-\frac{Q}{\Gamma}\right),\label{gcorr_terms1}\\
\ii \sigma Q &= c_s^2(r_0,z)\left(\ii k_r + \hat{g}_c
  \left.\frac{\p\ln{\rho}}{\p r}\right|_0\right)\delta v_x + c _s^2(r_0, z)\left( \left.\frac{\p\ln\rho}{\p
      z}\right|_0 \delta v_z + \frac{d\delta
    v_z}{dz}\right),\label{gcorr_terms2} \\
\ii \sigma \delta v_x & = \ii k_x W  -
\hat{g}_c\frac{1}{c_s^2(r_0,z)}\left.\frac{1}{\rho}\frac{\p P}{\p
  r}\right|_0Q - 2\Omega(r_0,z)\delta v_y,\label{gcorr_terms3}
\end{align}
where subscript $0$ denotes evaluation at $r=r_0$.

In the VGSB, the terms proportional to $\hat{g}_c$ are ignored. For a
power-law disk, these radial gradients are $O(r^{-1})$, and they
appear in comparison with terms of $O(k_x)$. The neglected terms 
therefore have a relative magnitude of $O(\epsilon/\khat)$. This is
small for thin disks ($\epsilon\ll1$) and/or small radial wavelengths
($\khat\gg 1$).

@@

\subsection{Spurious growth of adiabatic perturbations in the VGSB}\label{analytic_adia} 
An important caveat of the VGSB is that it cannot be used in the
adiabatic limit. We explain this by setting  $\beta\to\infty$ in
Eq. \ref{vertiso_gov_nondim}, which can then be written as 
% For adiabatic perturbations ($\gamma>1$, $\beta\to\infty$) we may
% simply use the Soilberg-Hoiland critera to assess stability
% (\S\ref{solberg}). Nevertheless, we briefly consider this case in the
% local framework to demonstrate its limitations. For $\beta\to\infty$
% the parameter $\chi = 1/\gamma$ and Eq. \ref{vertiso_gov_nondim} can
% be written as   
\begin{align}
  0 =\dd v_z^{\prime\prime} + \left(1 + \ii \epsilon q
    \hat{k}\right)\left(\ln\rho^{\prime}\delta v_z\right)^\prime
  +\left\{\hat{\sigma}^2\left(\frac{1}{\gamma}+\hat{k}^2\right) 
    -\left(\frac{\gamma-1}{\gamma}\right)\left[\ln\rho^{\prime\prime}+\hat{k}^2\left(1-\frac{\ii\epsilon  
          q}{\hat{k}}\right)\ln\rho^{\prime 2}\right]\right\}\delta v_z.\label{adia_iso3}
\end{align}
We multiply Eq. \ref{adia_iso3} by $\rho\delta v_z^*$ and
integrate vertically, assume boundary terms vanish when integrating by
parts, to obtain
\begin{align}
  \hat{\sigma}^2\left(\frac{1}{\gamma} +
    \hat{k}^2\right)\int_{\zhat_1}^{\zhat_2}\rho|\delta
  v_z|^2 d\zhat 
  =&  \left(\frac{\gamma-1}{\gamma}\right)
  \int_{\zhat_1}^{\zhat_2}\rho|\delta v_z^\prime|^2 d\zhat
  +\frac{1}{\gamma}\int_{\zhat_1}^{\zhat_2}\frac{1}{\rho}|(\rho\delta
  v_z)^\prime|^2 d\zhat\notag\\
&+
  \left(\frac{\gamma-1}{\gamma}\right)\hat{k}^2\left(1-\frac{\ii\epsilon
      q}{\hat{k}}\right) \int_{\zhat_1}^{\zhat_2}\rho\ln\rho^{\prime
    2}|\delta v_z|^2 d\zhat
+ \ii\epsilon q \hat{k}
  \int_{\zhat_1}^{\zhat_2}\ln\rho^\prime(\rho\delta v_z^*)^\prime
  \delta v_z d\zhat.\label{adia_integral}
\end{align}
When $q\equiv0$, all the terms on the right-hand-side (RHS) are real. Then
$\hat{\sigma}^2$ is real, and  $\hat{\sigma}^2>0$ if $\gamma>1$. As
expected, a sub-adiabatically stratified disk is stable in the absense
of vertical shear.

However, in the presence of vertical shear $q\neq0$,
Eq. \ref{adia_integral} suggests $\hat{\sigma}$ is complex for real
$\khat$, which implies the possibility of instability, regardless of
how much $\gamma$ deviates from unity. This is contradicts the second 
Solberg-Hoiland criterion (Eq. \ref{solberg2}), which states that
instability in the adiabatic limit requires the disk to be close to
neutrally stratified. %also need real freq^2 in adia limit 
This is because the VGSB retains the destabilizing effect of vertical shear
($\p_z\Omega$) from the global disk, but ignores the global radial disk
struture responsible for it. Nevertheless, we demonstrate below that
this inconsistency is unimportant for the vertical shear instability,
which occurs for $\beta\ll1$.

% To 
% avoid this, we must restrict our attention to short thermal relaxation
% timescales ($\beta\lesssim1$) and/or neutrally-stratified disks
% ($\gamma=\Gamma$) when $q\neq0$.

\subsection{Effect of global radial gradients}
% We may also reverse the above procedure by reinstating the `global
% correction' terms proportional to $\hat{g}_c$ in the linear VGSB
% equations. Of course, doing so implies the basic state is not uniform
% in $x$, which would be inconsistent with the full VGSB equations in
% equilibrium. Nevertheless, we illustrate in Fig. \ref{gcorr_compare} the effect of
% these global radial gradients on the growth rate of the fundamental
% VSI.

% The global correction terms do not change the qualitative dependence
% of the fundamental VSI growth rates on the thermal relaxation
% timescale. The difference is negligble for $\beta\lesssim 1$ (i.e.
% fast thermal relaxation). The critical relaxation timescale for the
% fundamental VSI is unaffected by these global radial radients.

% However, in the adiabatic limit $\beta\to\infty$ the linear VGSB
% equations give a positive growth rate as discussed in
% \S\ref{analytic_adia}, which is inconsistent with the second
% Solberg-Hoiland criterion (Eq. \ref{solberg2}). The correct behaviour
% ($\nu\to0$ as $\beta\to\infty$) is obtained with global correction.

In  Fig. \ref{gcorr_compare}, we demonstrate the effect of the global radial gradient terms
proportional to $\hat{g}_c$ in 
Eq. \ref{gcorr_terms1}---\ref{gcorr_terms3} by calculating the
fundamental VSI growth rates using three approaches. We compute growth rates from the dispersion
relation Eq. \ref{relax_disp}, which utilizes the VGSB;  from the
full linear problem (Eq. \ref{lin_mass}---\ref{lin_energy}) with
$\hat{g}_c=0$ (i.e. the linear VGSB equations); and from the full
linear problem with $\hat{g}_c=1$.

All three methods give similar behavior, and growth rates are in close
agreement for $\beta\lesssim 1$. Differences arise for
$\beta\gtrsim1$, and as $\beta\to\infty$ the VGSB framework gives a
(spurious) positive growth rate as expected from the discussion
above. Inclusion of the global radial gradient terms results in the
correct behavior  ($\nu\to0$ as $\beta\to\infty$).

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.0cm 0cm
  0cm]{figures/gcorr_compare} 
  \caption{Growth rate of the fundamental VSI mode as a function of
    the thermal relaxation time $\beta$. The `Analytical (VGSB)' curve
    is calculated from the dispersion relation Eq. \ref{relax_disp};
    the `Numerical (VGSB)' curve is obtained by numerically solving
    Eq. \ref{lin_mass}---\ref{lin_energy} with $\hat{g}_c=0$
    (which defines the linear VGSB equations); and the `Numerical
    (Full)' curve is obtained by numerically solving
    Eq. \ref{lin_mass}---\ref{lin_energy}  with $\hat{g}_c=1$, which
    accounts for the radial disk structure.  
% computed from the linear VGSB equations used in
%     the main text (dotted) and that including additional global
%     radial gradient terms in
%     Eq. \ref{gcorr_terms1}---\ref{gcorr_terms3} (solid). 
    The disk parameters are $\Gamma=1.011$,
    $\gamma=1.4$ and $(p,q,\epsilon)=(-1.5,-1,0.05)$. The perturbation
    wavenumber is $\khat=30$. 
    \label{gcorr_compare}}  
\end{figure}

Fig. \ref{gcorr_compare} shows that provided we consider $\beta\ll1$,
then the VGSB framework is an adequate approximation. It is
interesting to note that there is a thermal relaxation timescale that
maximizes the mode decay rate. Here, it is $\beta\simeq2$ or about
$0.3$ orbital periods. This is, in fact, consistent with Fig. 12 of
\cite{nelson13}.

### 2015-02-09 linear_iso.tex
major updating

\section{Analytical discussion of vertically isothermal
  disks in the framework of the vertically global shearing
  box}\label{analytical}  
We begin with a discussion of vertically isothermal disks
where $\Gamma=1$, then $c_s$ is a constant. We  
first examine isothermal perturbations ($\beta = 0$) before
considering the effect of finite thermal relaxation. 
In order to facilitate an analytical discussion, we introduce a number
of approximations below that allows a simple governing equation to be
derived and directly compared with previous works.

>>>>

\section{Analytical discussion of vertically isothermal
  disks in the framework of the vertically global shearing
  box}\label{analytical}  
We begin with a discussion of vertically isothermal disks
where $\Gamma=1$, then $c_s$ is a constant. We  
first examine isothermal perturbations ($\beta = 0$) before
considering the effect of finite thermal relaxation. 
In order to facilitate a fully analytic discussion, we introduce a number
of approximations below that allows a simple governing equation to be
derived and compared with previous works.

### 2015-02-09 numerical.tex
major updating

% mention mode selection -> focus on fundamental corrugation ->
% simulations eventually dominated by them 
% choice of kx is O(10), since nelson observe kxH \sim 30

@@

However, for $\khat\lesssim 10$, e.g. $\khat=1$ in 
Fig. \ref{lowfreq_eigenfunc}, the perturbation magnitude near the
boundaries begin to dominate. For even smaller $\khat\lesssim 1$ (not
shown) we find the perturbation is almost entirely concentrated at the
boundaries. Our radially-local approximation is likely invalid for 
these small wavenumber perturbations and we do not consider such modes
further.

@@

\subsection{Effect of thermal relaxation}
We now consider stably stratified disks ($\gamma > \Gamma$) with
thermal relaxation $\beta\in[10^{-3},1]$. Here we use a slightly  
different disk model with $\epsilon=0.05$ and $\gamma=1.4$, as 
adopted in simulations performed by \cite{nelson13}. Other parameters 
are unchanged from the previous section.

Fig. \ref{bcrit_compare1} shows the growth rate $\nu$ of the
fundamental VSI mode as a function of $\beta$ for $\khat=30$. This is
approximately the perturbation wavenumber observed in the fiducial
numerical simulation of \cite{nelson13}. We also plot the growth rates
obtained by numerically solving the dispersion relation 
Eq. \ref{relax_disp} and the upper limit to 
the thermal relaxation timescale $\beta_\mathrm{crit}$ given by
Eq. \ref{iso_vsi_cond}. For the present disk parameters
$\beta_\mathrm{crit}=0.125$.

We find good agreement between the numerical and analytical growth
rates, as well as the critical relaxation timescale. Growth rates are
insensitive to $\beta$ for $\beta\lesssim 0.05$, beyond which growth
rates rapidly decrease. For this disk model, \citeauthor{nelson13}
found a thermal relaxation timescale of $\beta\gtrsim 0.6$ stabilized
the VSI in non-linear hydrodynamic simulations. This is consistent
with our results.

@@

 \begin{figure}
   \includegraphics[width=\linewidth]{figures/bcrit_compare} 
   \caption{Growth rate of the fundamental VSI mode with $\khat=30$ as
     a function of the thermal relaxation timescale $\beta$. The disk
     parameters are $\Gamma=1.011$, $\gamma=1.4$, and 
     $(p,q,\epsilon)=(-1.5,-1,0.05)$. The numerical (analytical) growth
     rates are show as solid (dotted) lines, while the vertical line
     is the maximum thermal timescale $\beta_\mathrm{crit}$ for the
     fundamental mode according to Eq. \ref{iso_vsi_cond}. 
     \label{bcrit_compare1}}   
 \end{figure}

@@

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/eigenvectorW_beta0d01} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  0cm]{figures/eigenvectorW_beta0d1} 
  \caption{Numerically-calculated eigenfunction of the fundamental VSI
    mode in a nearly vertically isothermal disk ($\Gamma=1.011$) with
    $(p,q,\epsilon)=(-1.5,-1,0.05)$, evolved with $\gamma=1.4$ and a dimensionless 
    thermal relaxation timescale $\beta \equiv t_c\Omega_k = 0.01$
    (top) and $0.1$ (bottom). The
    perturbation radial wavenumber is $\khat=30$. The
    real (imaginary) part of $W$ is plotted as the solid (dotted)
    line.
    \label{relax_eigenW_num}}  
\end{figure}

\subsection{Critical thermal relaxation timescale}
% may need to explain why choose k=10
% bcrit independent of k for k>>1 
% for some parameter regimes (e.g. large q, large epsilon) 
% behavior at larger kx and finite bcool not understood. 
We test our estimate for the maximum thermal relaxation timescale 
$\beta_\mathrm{crit}$ for the fundamental
VSI (Eq. \ref{iso_vsi_cond}). We use the previous disk
parameters as a reference point and vary $q\in[-0.2,-1.2]$,
$\gamma\in[1.2,2.0]$, and $\epsilon\in[0.02,0.1]$
separately.

Here we use a wavenumber $\khat=10$.

% , in which case we find $\nu$
% changes sign as $\beta$ is increased from zero. We find that with
% $\khat=30$, as used in the previous section, when $\beta$ is not small $\nu$ did not reach zero but instead decreases
% rapidly after the expected value of $\beta_\mathrm{crit}$.  These
% modes were found to have large amplitudes near the boundaries.

%we suspect boundary conditions may play a role for

% For large $\khat$ \emph{and} finite
% thermal relaxation (e.g. $\khat=30$, $\beta\gtrsim 0.2$) we found
% $\nu$ did not reach zero, but still decreases rapidly after the
% expected $\beta_\mathrm{crit}$. These modes have large perturbations
% near the boundaries, which may play a role not accounted for in our
% analytical discussion.

The numerically-obtained $\beta_\mathrm{crit}$ is shown in
Fig. \ref{bcrit_compare} in comparison with Eq. \ref{iso_vsi_cond}.

Our numerical results generally agree with 
Eq. \ref{iso_vsi_cond}. The agreement improves with decreasing  $|q|$,
$\epsilon$ and increasing $\gamma$, i.e. weaker instability. 
There is a noticeable difference as $\gamma\to1$ because the
derivation of $\beta_\mathrm{crit}$ assumed $\gamma>1$.

@@

% \subsubsection{$\gamma=\Gamma \neq 1$, $N_z^2\equiv 0$}
% Consider first the case $\gamma=1.4$ so that there is no vertical
% entropy gradient. We plot in Fig. the growth rates for the
% fundamental VSI for $\beta\in[10^{-2},10^{2}]$. We find growth rates
% do not reach zero, implying instability regardless of the thermal
% relaxation timescale if $\gamma=\Gamma$. This result is consistent
% with \cite{nelson13}.  Note that for $\beta\to\infty$ (adiabatic
% perturbations) and $\gamma=\Gamma$, the disk can be unstable according
% to the Solberg-Hoiland criterion  Eq. \ref{solberg2}.

% Growth rates for $\beta = 10^{-2}$ and
% $\beta=10^{2}$ behave similarly. This is not surprising because
% the linearized energy equation (Eq. \ref{lin_energy}) implies that 
% \begin{align}
%   W = \frac{\left(\Gamma^{-1} -
%       \ii\hat{\sigma}\beta\right)}{\left(1-\ii\hat{\sigma}\beta\right)}
%   Q \text{ for } \gamma\equiv \Gamma.   
% \end{align}
% Thus, $W \to Q/\Gamma$ as $\beta\to0$ and $W\to Q$ as
% $\beta\to\infty$. Since $\Gamma=O(1)$, these two limits are expected
% to produce qualitatively similar results. Interestingly,
% Fig. \ref{growth_vnoniso1} show that growth rates 
% maximize for $\beta=O(1)$. However, all growth rates are similar to
% order-of-magnitude for $0.1\lesssim\khat\lesssim 100$.

% \begin{figure}
%   \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
%   0cm]{figures/growth_vnoniso1} 
%   \caption{Numerically-determined growth rates of the fundamental VSI
%     mode in neutrally-buoyant ($N_z^2\equiv0$), vertically non-isothermal disks
%     ($\Gamma=\gamma=1.4$), with $(p,q,\epsilon)=(0,-1,0.05)$, for
%     a range of thermal relaxation timescales 
%     $\beta$. \label{growth_vnoniso1}}     
% \end{figure}

% \subsubsection{$\gamma\neq\Gamma \neq 1$, $N_z^2\geq0$}
% %gamma = 2.0, Gamma = 1.4, vary bcool 
% Finally, we introduce buoyancy in a vertically non-isothermal disk by
% setting $\gamma=2$. This scenario was not considered by  
% \cite{nelson13}, but is similar in spirit with vertically isothermal 
% disks evolved with $\gamma>1$, as considered in the previous section. 
% Fig. \ref{growth_vnoniso2} plots the fundamental VSI growth rates for
% this case. We observe the same trend as for vertically isothermal
% disks with $N_z^2\geq 0$ (Fig. \ref{relax_growth_num}), namely growth
% rates reach zero at finite $\khat$. From Fig.\ref{growth_vnoniso2} we
% expect the VSI to be strongly stabilized even for small, finite
% thermal relaxation times.

% \begin{figure}
%   \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
%   0cm]{figures/growth_vnoniso2}
%   \caption{Same as Fig. \ref{growth_vnoniso1}, but with $\gamma=2$ so that
%     $N_z^2\geq0$. \label{growth_vnoniso2}}
% \end{figure}

>>>>

% mention mode selection -> focus on fundamental corrugation ->
% simulations eventually dominated by them 
% choice of kx is O(10), since nelson observe kxH \sim 30
%not possible to survey all modes/kx/disk parameters
%focus on effect of thermal relaxation

@@

However, for $\khat\lesssim 10$, e.g. $\khat=1$ in 
Fig. \ref{lowfreq_eigenfunc}, the perturbation magnitude near the
boundaries begin to dominate. For even smaller $\khat\lesssim 1$ (not
shown) we find the perturbation is almost entirely concentrated at the
boundaries. Our radially-local approximation is not valid for 
these small wavenumber perturbations and we do not consider such modes
further.

@@

\subsection{Effect of thermal relaxation}
We now consider stably stratified disks ($\gamma > \Gamma$) with
thermal relaxation. Here, we use a slightly  
different disk model with $\epsilon=0.05$ and $\gamma=1.4$, as 
adopted in simulations performed by \cite{nelson13}. Other parameters 
are unchanged from the previous section.

Fig. \ref{bcrit_compare1} shows the growth rate $\nu$ of the
fundamental VSI as a function of $\beta$ for $\khat\in[1,10^2]$. We
also plot the critical thermal relaxation timescale
$\beta_\mathrm{crit}$ given by Eq. \ref{iso_vsi_cond}. For the present
disk parameters $\beta_\mathrm{crit}=0.125$.

For $5\lesssim\khat\lesssim30$, we find $\beta_\mathrm{crit}$ provides an
accurate prediction of the upper limit to the thermal relaxation 
timescale for  the fundamental mode. For smaller $\khat$,
$\beta<\beta_\mathrm{crit}$ is a sufficient condition for
instability. For larger $\khat$, we find $\nu$ does not reach zero but
rapidly decreases as $\beta$ approaches and increases beyond
$\beta_\mathrm{crit}$. In the latter case we find the perturbation is
concentrated at the vertical boundaries, which may play a role not
accounted by our analytical discussion.

Fig. \ref{bcrit_compare1} shows that 
$\beta_\mathrm{crit}$ is a good indication for the maximum
thermal relaxation timescale to permit the fundamental VSI. For
$\beta\lesssim O(10^{-2})$, growth rates are insensitive to thermal
relaxation. For this disk model, \citeauthor{nelson13}
found a thermal relaxation timescale of $\beta\gtrsim 0.6$ stabilized
the VSI in non-linear hydrodynamic simulations. This is consistent
with our results.

@@

 \begin{figure}
   \includegraphics[width=\linewidth]{figures/bcrit_compare} 
   \caption{Growth rate of the fundamental VSI as
     a function of the thermal relaxation timescale $\beta$. The disk
     parameters are $\Gamma=1.011$, $\gamma=1.4$, and 
     $(p,q,\epsilon)=(-1.5,-1,0.05)$. The vertical line
     is the critical thermal timescale $\beta_\mathrm{crit}$ for the
     fundamental mode according to Eq. \ref{iso_vsi_cond}. 
     \label{bcrit_compare1}}   
 \end{figure}

@@

%approximately the perturbation wavenumber observed in the fiducial
%numerical simulation of \cite{nelson13}.

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/eigenvectorW_beta0d01} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  0cm]{figures/eigenvectorW_beta0d1} 
  \caption{Numerically-calculated eigenfunction of the fundamental VSI
    mode in a nearly vertically isothermal disk ($\Gamma=1.011$) with
    $(p,q,\epsilon)=(-1.5,-1,0.05)$, evolved with $\gamma=1.4$ and a dimensionless 
    thermal relaxation timescale $\beta = 0.01$
    (top) and $\beta=0.1$ (bottom). The
    perturbation radial wavenumber is $\khat=30$. The
    real (imaginary) part of $W$ is plotted as the solid (dotted)
    line.
    \label{relax_eigenW_num}}  
\end{figure}

\subsection{Critical thermal relaxation timescale}
% may need to explain why choose k=10
% bcrit independent of k for k>>1 
% for some parameter regimes (e.g. large q, large epsilon) 
% behavior at larger kx and finite bcool not understood. 
We test the dependence of the critical thermal relaxation timescale 
$\beta_\mathrm{crit}$ for the fundamental VSI on disk parameters
(Eq. \ref{iso_vsi_cond}).  We use the previous case study as a
reference point and vary $q\in[-0.2,-1.2]$, 
$\gamma\in[1.2,2.0]$, and $\epsilon\in[0.02,0.1]$
separately. Here, we use a wavenumber $\khat=10$, for which we
find growth rates reach zero for the range of disk parameters
considered, so that a numerical $\beta_\mathrm{crit}$ can be defined precisely.

The numerically-obtained $\beta_\mathrm{crit}$ is shown in
Fig. \ref{bcrit_compare} in comparison with Eq. \ref{iso_vsi_cond}.
Our numerical results generally agree with Eq. \ref{iso_vsi_cond}. The
agreement improves with decreasing  $|q|$,  $\epsilon$ and increasing
$\gamma$, i.e. weaker instability, although there is a noticeable difference
as $\gamma\to1$ because the derivation of $\beta_\mathrm{crit}$
assumed $\gamma>1$.

Fig. \ref{bcrit_compare}, together with the previous section
(Fig. \ref{bcrit_compare1}), shows that $\beta_\mathrm{crit}$ as given
by Eq. \ref{iso_vsi_cond} gives a measure of the characteristic
thermal timescale below which perturbations are effectively isothermal
so that the fundamental VSI operates.

% From the previous section (Fig. \ref{relax_eigenW_num}) we
% expect that for other wavenumbers,  still gives a characteristic
% thermal timescale below which the perturbations are effectively
% isothermal.

@@

Fig. \ref{gcorr_compare_vnoniso} plots the fundamental VSI growth
rates as a function of $\beta$ for $\gamma\in[1.4,2.5]$. For
comparison with \cite{nelson13} we consider 
$\beta\in[10^{-2},10^2]$. In agreement with \citeauthor{nelson13}, in
the neutrally-stratified case $\gamma=\Gamma$ the disk is unstable
even for $\beta\gg 1$. (Note that in the adiabatic limit
$\beta\to\infty$ this disk can be unstable according to the
Solberg-Hoiland criterion, Eq. \ref{solberg2}.)

For $\gamma>\Gamma$, i.e. stably stratified disks,
Fig. \ref{gcorr_compare_vnoniso} shows that introducing finite thermal
relaxation rapidly stabilizes the fundamental VSI mode. This is
similar to that observed for nearly vertically isothermal disks
(Fig. \ref{bcrit_compare1}). For $\gamma=1.7,\,2.0,\,2.5$, growth
rates reach zero for  $\beta\simeq0.17,\,0.083,\,0.045$, respectively.
Interestingly, these values are equal to
$\epsilon|q|/(\gamma-\Gamma)$. This suggests that the critial thermal
relaxation timescale for vertically non-isothermal disks can also be
estimated by Eq. \ref{iso_vsi_cond} but with $\gamma-1$ replaced by
$\gamma-\Gamma$.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  0cm]{figures/gcorr_compare_vnoniso}
  \caption{Growth rate of the fundamental VSI mode as a function of
    the thermal relaxation timescale $\beta$, in vertically
    non-isothermal disks with $\Gamma=1.4$ and
    $\gamma\in[1.4,2.5]$. The disk is neutrally
    stratified for $\gamma=1.4$ and stably stratified for
    $\gamma>1.4$. Other disk parameters are
    $(p,q,\epsilon)=(0,-1,0.05)$ and the perturbation wavenumber is
    $\khat=30$.  
    \label{gcorr_compare_vnoniso}}
\end{figure}
